\begin{fulltable}{|c|c|c|}{Common OTP behaviours}{tab:otp_behaviours}
\hline
Module & Description & Callbacks \\
\hline
\hline
\multirow{4}{*}{gen\_server.erl}
& Generic server behaviour. Contains a state  & init,        \\
& that is manipulated by calls (that require  & handle\_call,\\
& a reply) and casts (that do not wait for a  & handle\_cast,\\
& reply).                                     & terminate    \\
\hline
\multirow{5}{*}{gen\_fsm.erl}
& Finite state machine behaviour. A finite    & init, State1,\\
& number of states exist along with the       & State2, ..., \\
& messages each state accepts, the replies    & StateN,      \\
& that are sent and the state change that     & terminate    \\
& may follow.                                 &              \\ 
\hline
\multirow{3}{*}{gen\_event.erl}
& Generic event handler. Event handlers       & init,         \\
& register in a central event manager and     & handle\_event,\\
& are notified for any event that arrives.    & terminate     \\
\hline
\multirow{4}{*}{application.erl}
& \er\ application. An application is a       &       \\
& collection of modules that implement some   & start,\\
& specific functionality and can be started   & stop  \\
& and stopped as a whole.                     &       \\
\hline
\multirow{4}{*}{supervisor.erl}
& A process which supervises other processes  &     \\
& called child processes. A child process can & init\\
& either be another supervisor or a worker    &     \\
& process.                                    &     \\
\hline
\end{fulltable}
