\chapter{Related and Further work}

\section{Related work}
\label{sct:related_work}

There is almost no other known cases where tools need to make checks
like the ones implemented for \er's \emph{behaviours}, either in their
generic usage or in the detection of races through
them. Object-oriented languages that use the equivalent \emph{abstract
  classes}, \emph{virtual methods} and \emph{interfaces} have static
typing (C++, Java, OCaml, ...) which ensures the fitting of
implementations in every case.

On the subject of intersection types in a dynamically typed
programming language, only DRuby is a known analog and is presented
below. Other related work has been centered around Dylan and
JavaScript and is briefly mentioned as well.

Finally, the changes in the refinement procedure are related with a
formal approach to control flow analysis, also described below

\subsection{Diamondback Ruby (DRuby)}

Diamondback Ruby (DRuby)\cite{druby} is a recent tool that blends
Ruby's dynamic type system with a static typing discipline. It uses a
similar approach as \dr\ generating constraints for the variables then
applying a set of rewrite rules exhaustively. Intersection types are
included from the beginning in it's type system, but it cannot infer
them automatically. As a result they need to be annotated by the
developer placing them on the same level as \dr's contracts
(\emph{specs}).

\subsection{Dylan and JavaScript}

Dylan is a dynamically typed object-centered programming language
inspired by Common Lisp and ALGOL. In a recent publication Mehnert
proposed an extension providing function types and parametric
polymorphism to the language\cite{dylan}. The function types
specialize from the previous generic ones but do not include
intersections. Powerful parametric polymorphism is provided though.

JavaScript is the main scripting language for Web browsers, and it is
essential to modern Web applications. Applying type analysis to
JavaScript is a subtle business because, like most other scripting
languages, JavaScript has a weak, dynamic typing discipline which
resolves many representation mismatches by silent type conversions. In
their publication\cite{javascript} Jensen, M\o ller and Thiemann
develop such a type analyzer which like \dr\ is fully automatic
\emph{but} is designed for \emph{soundness} with regard to the absense
of certain errors.

\dr's success typings approach is mentioned in both these attempts to
provide static typing and discrepancy detection to dynamically typed
languages.

\subsection{Refinement using control flow analysis}

In their publication, Palsberg and Pavlopoulou \cite{polyvariant}
propose an approach to control flow analysis that goes one step
further than the changes proposed in this thesis (Section
\ref{sct:intersect_refinement}). Instead of using the types of the
arguments as they are presented in the various calls, a computation is
performed to obtain ``covers'' from them and use these covers instead
in the polyvariant refinement. For this idea to be applicable we need
to figure out if the actual calls can be reduced to some elementary
sets such that each call is exactly covered by some of them.

\section{Further work}
\label{sct:further_work}

\subsection{Behaviours}
\subsubsection{Automatic bypass of API for race detection}

The ``bypass'' mechanism was designed to be extensible, allowing other
behaviour API's to be connected with the respective callbacks. Anyone
with a better understanding of the other behaviours may document the
rest OTP's behaviours easily by adding code in
\emph{dialyzer\_behaviours} module. It would be even better if this
connection was tranferred in each behaviour's file, as a new attribute
or as an extension on the \emph{callback} attribute that was
introduced in this thesis.

\subsection{Intersections}

\subsubsection{Negative types}
\label{sct:negative_types}

The next logical extension to the type system would be negative
types. Examples would be \emph{``any term except the integer 42''} or
\emph{``any atom except a''}. With this infrastructure, when
\dr\ generates disjunctive lists it will be able to eliminate the
types already covered in previous clauses in the following ones. Thus
a ``catch-all'' will not have type \any\ but \emph{``anything but X, Y
  and Z''} where \emph{X, Y} and \emph{Z} will be the types already
covered by previous clauses.

\subsubsection{Tighter coupling between type and code}

Using normal form and sorting on the function's type has an impact on
the relation between the type of the function and the actual code that
generated it. Even though it was easy to find the cause of all the
warnings emitted when intersection types were used, it might be better
to narrow down a warning to a particular clause instead of the generic
pointer to the first line of the function.

\subsubsection{Better refinement}

In the current form, if a lot of calls are present in the code the
refinement analysis might fail due to the added effect of the
disjunctive refinement constraint. The ideas presented in Palsberg and
Pavlopoulou \cite{polyvariant} might reduce the elements of the
constraint in such cases, maintaining strictness.
